Some forms of uncertainty may suppress the evolution of social learning

### 2022-04-07 ms.tex
started journal article with updated model

@@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{{Figures/SL_over_u_lowpayoff=0.1.pdf}}
  \caption{}
  \label{fig:}
  \end{subfigure}

@@

  \caption{}
  \label{fig:}
\end{figure}

>>>>

We developed an agent-based model of a society of $N$ individuals who each must decide
which of $B$ behaviors to perform at each time step. Each behavior is a ``bandit'',
a common modelling and experimental approach for representing behaviors with
probabilistic payoffs~\cite{SuttonBartoBook, McElreath2005,Rendell2010,Schulz2019}. %Bandits are like  slot machines in a casino~\ref{fig:behaviorSelection}.
Each behavior $b$ yields a payoff of 1 with probability $\pi_b$ and a payoff of zero
otherwise; $\pi_b$ is therefore also the expected payoff of behavior $b$. Agents must
decide which behavior to perform at each time step. To do this, agents 
use an explore-exploit strategy to sometimes try the most profitable behavior
they know about, and other times try alternatives that may pay off more reliably.

We operationalized uncertainty in four different ways: 
(1) payoff ambiguity, $A_\pi$, which measures the difference
between the optimal expected payoff behavior $\pi_{high}$ and the expected payoff
of the other behaviors, $\pi_{low}$; 
(2) environmental variability, $u$, the probability the optimal behavior changes from one generation to another; 
(3) the number of possible behaviors, $B$; and 
(4) the lifespan, or time steps per generation, $M$.

We ran the simulation for $T=1000$ time steps. At
the end of each generation agents reproduce and then die off. 
Those selected to reproduce
pass on their boolean social learning trait, $s$, without mutation,
which specifes whether agents learn socially. We developed a series of
computational analyses where we systematically vary the uncertainty
variables and observe the average prevalence of social learning trait $s$
across agents and trials, which is our primary outcome measure, 
denoted $\langle s \rangle$.

\subsection{Agents and learning}

In each time step, $N$ agents---autonomous problem solvers---select 
which behavior to perform based on their running tally of mean payoffs for 
each available behavior. %information (Figure~\ref{fig:behaviorSelection}). 
Agents track the mean payoff of each behavior $b$, denoted $\bar\pi_b$, and a \emph{count} of how many times it has performed each behavior, denoted $c_b$. 
Agents' beliefs about mean payoffs are updated from $\bar\pi_b$ to $\bar\pi_b'$ using exponential weighted
averaging, $\bar\pi_b' = \bar\pi_b + \frac{\mathrm{Bandit_b(0, 1)} - \bar\pi_b}{c_b'}$, where
$\mathrm{Bandit_b(0, 1)}$ is 0 or 1 depending on the result of the bandit draw for behavior $b$. 
$\bar\pi_b$ is initialized to 0 for all $b$ at model initialization for
all agents, and after each generation for individual learner agents. Social
learning agents' $\bar\pi_b$ is initialized as that of its teacher selected
through performance-biased oblique learning.

Social learning in this model is oblique, which here means
any agent from the previous generation, including an agent's parent,
could be an agent's teacher. 
Agents learn from exactly one teacher, selected as follows:
First, a learner agent from a new generation selects $N_T$ teachers at random.
The best-performing of the $N_T$ becomes a new agent's teacher, with ties broken
randomly. At the end of each generation, all agents die off with some selected
randomly for reproduction, weighted by accumulated payoffs within the round.

Following Luce's choice axiom \cite{Luce1959}, an agent using asocial information selects a behavior at random, with the probability of selecting any particular behavior weighted by the softmax function applied to that behavior's observed mean payoff relative to all 
mean payoffs: 
$\Pr(\text{choose behavior }b) = 
  \frac{\exp(\bar\pi_b / \tau) }{ 
\sum_{b=1}^B \exp(\bar\pi_b / \tau)}$.
The softmax method is a biologically plausible model~\cite{Schulz2019} that enables agents to explore alternative behaviors sometimes and exploit the best observed behavior other times.  The parameter $\tau$ specifies
how frequently alternative behaviors are explored---higher $\tau$ means more
frequent exploration. We set $\tau = 0.1$ for all computational analyses 
presented in the main paper. We (WILL) perform(ED) sensitivity analyses 
and found a weak dependence on $\tau$ that does not affect our main conclusions.

\subsection{Modeling uncertainty}

In our model uncertainty is a tunable consequence of four environmental features:
%that structure the behavior of agents.  
% Many factors contribute to the 
% uncertainty involved in both social and asocial learning. 
%We model uncertainty in four distinct ways: 
\begin{enumerate}
    \item We vary the frequency at which the optimal behavior changes, which we call \emph{environmental variability}, $u$. At the start of each generation, with probability $u$, a new behavior is assigned a payoff of $\pi_{high}$ while all other behaviors are assigned a payoff of $\pi_{low}$. Otherwise, the same behavior remains optimal across generations. 
    \item We vary the latent expected payoffs yielded by the bandits. For simplicity,
      we assume that in any given environmental state, there is one optimal behavior that
      yields an expected payoff of $\pi_{high}$, while all other behaviors yield a 
      payoff of $\pi_{low} < \pi_{high}$. 
      The difference between these quantities is the \emph{payoff ambiguity}, 
    \begin{equation}
      A_\pi = 1 - (\pi_{high} - \pi_{low}),
      \label{eq:payoffAmbiguity}
    \end{equation}
    which is maximized when the expected payoffs are close to equal and minimized
    when they differ greatly. For simplicity we will set $\pi_{high} = 0.9$
    for all analyses and vary $\pi_{low}$ in order to vary $A_\pi$.
    \item We vary the total \emph{number of available behaviors}, $B$, which is a
      source of uncertainty since agents are less likely to know which behavior yields high payoffs.  %risk, as it becomes increasingly costly to experiment with alternatives. \cm{is this risk if it reduces variance of outcomes? it sounds like it's just bad for there to be more 0 payoff options}
    \item Finally, we vary number of behavioral events per generation, $M$. This can be viewed as the \emph{effective lifespan} of an agent.  Decreasing this lifespan effectively increases the importance of each event for acquiring payoffs. When lifespans are short ($M$ is small), agents experience greater uncertainty about which behavior is optimal given the fewer learning opportunities.
\end{enumerate}

\subsection{Dynamics and evolution}

\subsection{Computational analyses}

We manipulated environmental uncertainty parameters described above, $u$,
$\pi_{low}$, $B$, and $M$, to examine their effects on our
main outcome measure, the prevalence of social learning $s$. For each parameter setting
in our analysis we calculated the average value of $s$ at the final time step
of $T=1000$ across 100 runs for each combination of uncertainty parameter values.
The model

@@

To analyze the effect of various forms of uncertainty on social learning 
prevalence, we analyzed the outcomes of several model runs initialized with
different uncertainty parameters $u$, $A_\pi$ (via $\pi_{low}$), 
and $M$ (AND SOON $B$???). Specifically, for each parameter setting over
$\pi_{low} \in \{0.1, 0.5, 0.8\}$ and $M \in \{1, 2, 4\}$, we varied 
$u \in \{0.0, 0.2, \ldots, 1.0\}$.
Our analysis confirmed existing theoretical predictions that social learning
is valuable only when environmental uncertainty is sufficiently small. However
our analysis also revealed that environmental uncertainty is not the only 
relevant factor. Payoff ambiguity can alter selection selection pressure for or
against social learning, depending on the number of time steps per generation.
We tested three values of $\pi_{low} \in \{0.1, 0.5, 0.8\}$, representing
low, moderate, and high values of $A_\pi$.

When $\pi_{low} = 0.1$, $M$ had little to no effect (Figure~\ref{fig:results_pilow_0p1}).
For all three values of $M$, social learning decreases from 1 dramatically over moderate
values of $u$, going to 0 by $u=0.6$. When $\pi_{low}=0.5$, however, the 
transition from $\langle s \rangle = 1$ to $\langle s \rangle = 0$ is not as
sharp, and it is slower for $M=1$ compared to $M=2$ and $M=4$. When $\pi_{low} = 0.8$
there appears to be a weak dependence of social learning on $u$ across $M$,
but this trend is mostly dominated by noise.
(EXPAND THIS---PROBABLY NEED B > 2, WHAT ELSE?)

(EXPLAIN WHY)

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.5\textwidth}
    \centering
    \includegraphics[width=\textwidth]{{Figures/SL_over_u_lowpayoff=0.1.pdf}}
  \caption{}
  \label{fig:results_pilow_0p1}
  \end{subfigure}

@@

  \caption{Dependence of Social Learning ($\langle s \rangle$) on environmental
  variability $u$ for several values of low payoff $\pi_{low}$ and steps per
  round, $M$. $\langle s \rangle$ decreases monotonically for $\pi_{low} = 0.1, 0.5$,
but noise dominates the results for $\pi_{low}=0.8$. When $\pi_{low} = 0.5$, 
the sigmoid is not as sharp as for $\pi_{low}=0.1$, meaning that social learning
is reduced for smaller $u$ and increased for larger $u$.}
  \label{fig:results}
\end{figure}
